---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
options(digits = 4, width = 120)
```

# datadict-cmd: Utilities to run the R package [datadict](https://github.com/epicentre-msf/datadict) from the command line

### Installing required R packages

From command line:
```
Rscript R/_install_deps.R
```

Or, from R:
```{r, eval=FALSE}
install.packages(c("readxl", "remotes"))
remotes::install_github("epicentre-msf/datadict")
```


### Check if data dictionary is valid

```
Rscript R/cmd_valid_dict.R [dict] [verbose]
```

#### Arguments:

Note that from command line arguments are currently unnamed and so must be specified in order (this can be improved in future).

1. `dict`: path to data dictionary file (must be .xlsx)
2. `verbose`: TRUE/FALSE indicating whether to give warnings describing the checks that have failed (if any). Optional, defaults to TRUE.

#### Outputs:

`TRUE` if all checks pass, `FALSE` if any checks fail. If `verbose = TRUE` and any checks fail, will also return description of checks that have failed.

#### Examples:

Path to valid dictionary, `verbose` unspecified (defaults to TRUE)
```
$ Rscript R/cmd_valid_dict.R data/dict_valid.xlsx
[1] TRUE
```

Path to valid dictionary, `verbose = FALSE`
```
$ Rscript R/cmd_valid_dict.R data/dict_valid.xlsx FALSE
[1] TRUE
```

Path to nonvalid dictionary, `verbose` unspecified (defaults to TRUE)
```
$ Rscript R/cmd_valid_dict.R data/dict_nonvalid.xlsx
[1] FALSE
Message d'avis :
- Missing values in column(s): "type"
- Duplicated values in column `variable_name`: "source_water" 
```

Path to nonvalid dictionary, `verbose = FALSE`
```
$ Rscript R/cmd_valid_dict.R data/dict_nonvalid.xlsx FALSE
[1] FALSE
```


### Check if dataset is valid (i.e. corresponds to data dictionary)

```
Rscript R/cmd_valid_data.R [dict] [data] [format_coded] [verbose]
```

#### Arguments:

1. `data`: path to dataset file (must be .xlsx), only first sheet is read
2. `dict`: path to data dictionary file (must be .xlsx)
3. `format_coded`: Are Coded-list type variables encoded as raw values ("value") or labels ("label") within the dataset. E.g. Variable `sex` might coded as 0/1 ("value") or "Male"/"Female" ("label"). Defaults to "label". Must be specified by user uploading the data.
4. `verbose`: TRUE/FALSE indicating whether to give warnings describing the checks that have failed (if any). Optional, defaults to TRUE.

#### Outputs:

`TRUE` if all checks pass, `FALSE` if any checks fail. If `verbose = TRUE` and any checks fail, will also return description of checks that have failed. If dictionary does not pass all checks will fail with error.

#### Examples:

Path to valid dataset, `verbose` unspecified (defaults to TRUE)
```
$ Rscript R/cmd_valid_data.R data/data_valid.xlsx data/dict_valid.xlsx
[1] TRUE
```

Path to nonvalid dataset, `verbose` unspecified (defaults to TRUE)
```
$ Rscript R/cmd_valid_data.R data/data_nonvalid.xlsx data/dict_valid.xlsx
[1] FALSE
Message d'avis :
- Columns defined in `dict` but not present in `data`: "ilness_other"
- Variables of type 'Numeric' contain nonvalid values: "age_years" 
```

Path to nonvalid dataset, set `verbose` to FALSE
```
$ Rscript R/cmd_valid_data.R data/data_nonvalid.xlsx data/dict_valid.xlsx label FALSE
[1] FALSE
```

Path to valid dataset, but set `format_coded` to "value" when in fact the format in the dataset is "label"
```
$ Rscript R/cmd_valid_data.R data/data_valid.xlsx data/dict_valid.xlsx value
[1] FALSE
Message d'avis :
- Variables of type 'Coded list' contain nonvalid values: "location", "cluster", "source_water", "sex", "age_under_one", "arrived", "departed", "born", "died", "illness", "oedema", "source_water_other", "cause_death", "cause_death_other", "ilness_other" 
```

Path to valid dataset, but dictionary is nonvalid
```
$ Rscript R/cmd_valid_data.R data/data_valid.xlsx data/dict_nonvalid.xlsx
Erreur : Dictionary does not pass all checks
Exécution arrêtée
```


### Check *k* anonymity, with manual specification of indirect identifiers

```
Rscript R/cmd_k_anonymity.R [data] [vars]
```

#### Arguments:

1. `data`: path to dataset file (must be .xlsx), only first sheet is read
2. `vars`: comma-separated list of relevant variables

#### Outputs:

Integer, the observed minimum value of *k* in the dataset. If this value is
greater than or equal to the pre-specified *k* anonymity threshold for the
project, then the data is sufficiently pseudonymized. If the observed value of
*k* is lower than the threshold, further pseudonymization is required.

#### Examples:

Assuming a pre-specified *k* of 5, the example below is sufficiently pseudonymized
```
$ Rscript R/cmd_k_anonymity.R data/data_valid.xlsx location,cluster,sex
[1] 37
```

Assuming a pre-specified *k* of 5, the example below is not sufficiently pseudonymized
```
$ Rscript R/cmd_k_anonymity.R data/data_valid.xlsx location,cluster,sex,source_water
[1] 1
```

Specify variable that doesn't exist in the dataset
```
$ Rscript R/cmd_k_anonymity.R data/data_valid.xlsx location,var_doesnt_exit
Erreur : The following variables do no exist in the dataset: "var_doesnt_exit"
Exécution arrêtée
```


### Check *k* anonymity, pulling indirect identifiers from data dictionary

```
Rscript R/cmd_k_anonymity_dict.R [data] [dict]
```

#### Arguments:

1. `data`: path to dataset file (must be .xlsx), only first sheet is read
2. `dict`: path to data dictionary file (must be .xlsx)

#### Outputs:

Integer, the observed minimum value of *k* in the dataset. If this value is
greater than or equal to the pre-specified *k* anonymity threshold for the
project, then the data is sufficiently pseudonymized. If the observed value of
*k* is lower than the threshold, further pseudonymization is required.

#### Examples:

Assuming a pre-specified *k* of 5, the example below is sufficiently pseudonymized
```
$ Rscript R/cmd_k_anonymity_dict.R data/data_valid.xlsx data/dict_valid.xlsx
[1] 37
```
